\todo{New title} Concretizing the Process Hitting into Biological Regulatory Networks

### 2013-04-25 parts/infer-param.tex
discussions on interval parameters & unsigned edges
start refactoring section 4.3

\begin{property}[Extreme values assumption]
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{extreme values assumption} if and only if:
\label{prop:param_enum_extreme}
\begin{align*}
  \forall b \in \Gamma, \GRNreg{b} \neq \emptyset \Longrightarrow \exists \omega \subset \GRNreg{b}, 0 \in K_{b,\omega} \wedge \exists \omega' \subset \GRNreg{b}, l_b \in K_{b,\omega'}
\end{align*}
\end{property}

\begin{property}[Activity assumption]
\label{prop:param_enum_activity}
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{activity assumption} if and only if:
\begin{align*}
  \forall b \in \Gamma, \forall a \in \GRNreg{b}, \exists \omega \subset \GRNreg{b}, K_{b,\omega} \neq K_{b,\omega \cup \{ a \}}
\end{align*}
\end{property}

\begin{property}[Monotonicity assumption]
\label{prop:param_enum_monotonicity}
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{monotonicity assumption} if and only if:
\begin{align*}
  \forall b \in \Gamma,
  \forall A^+ \subset \{ a \in \Gamma \mid \GRNedge{a}{+}{t}{b} \in E_+ \}&,
  \forall A^- \subset \{ a \in \Gamma \mid \GRNedge{a}{-}{t}{b} \in E_- \},\\
  K_{b,\omega \cup A^-} & \leqsegm K_{b,\omega \cup A^+}
\end{align*}
\end{property}

@@

Answer Set Programming (ASP) \cite{Baral03} has been chosen to address the enumeration of all admissible parametrizations.
The motivations are following:
\begin{itemize}
  \item ASP efficiently tackles the inherent complexity of the models we use, thus allowing a fast execution of the formal tools defined in this paper,
  \item it is convenient to enumerate a large set of possible answers,
  \item it allows us to easily constrain the answers according to some properties.
\end{itemize}
We now synthesize some key points to better make the reader understand our ASP implementation with the enumeration example.

All information describing the studied model (PH and inferred IG \& parameters) are expressed in ASP using facts.
For functional purposes, we assign a unique label to each couple $A,B$ of activators and inhibitors of a given component, and in the following we note $K^p_{a,A,B}$ the parameter of component $a$ whose regulators $A,B$ are assigned to the label $p$.
Then, to state the existence of a parameter $K^\ti{p}_{\ti{a},A,B}$, we use an atom named \aspil{param\_label} in the following fact:
\asp{param\_label(\ti{a}, \ti{p}).}

Defining a set in ASP is equivalent to defining the rule for belonging to this set.
For example, we define an atom \aspil{param\_act} that describes the set of active regulators of a given a parameter.
%of component \ti{a} and label \ti{p} (\ie the set $A$ of a parameter $K^\ti{p}_{\ti{a},A,B}$).
Describing the activators of $K^\ti{p}_{\ti{a},\{\ti{b},\ti{c}\},\{\ti{d}\}}$ gives:
\asp{param\_act(\ti{a}, \ti{p}, \ti{b}).
\item[] param\_act(\ti{a}, \ti{p}, \ti{c}).}
The absence of such a fact involving \ti{d} with label \ti{p} indicates that \ti{d} is an inhibitor in the configuration of regulators related to this parameter.

Rules allow more detailed declarations than facts as they have a body (right-hand part below) containing constraints and allowing to use variables, while facts only have a head (left-hand part).
For instance, in order to define the set of expression levels of a component, we declare:
\asp{component\_levels(X, 0..M) :- component(X, M).}
where the \aspil{component(X, M)} atom stands for the existence of a component \aspil{X} with a maximum level \aspil{M}.
Considering this declaration, any possible answer for the atom \aspil{component\_levels} will be found by binding all possible values of its two terms with all existing \aspil{component} facts: the existence of an answer \aspil{component\_levels(\ti{a}, \ti{k})} will depend on the existence of a term \ti{a}, which is bound with \aspil{X}, and an integer~\ti{k}, constrained by: $0 \leq \ti{k} \leq \aspil{M}$.

Cardinalities are convenient to enumerate all possible parametrizations by creating multiple answer sets.
A cardinality (denoted hereafter with curly brackets) gives any number of possible answers for some atoms between a lower and upper bounds.
For example,
\asp{1 \{ param(X, P, I) : component\_levels(X, I) \} :-
\item[] ~~~~~~param\_label(X, P), not infered\_param(X, P).}
where \aspil{param(X, P, I)} stands for: $\aspil{I} \in K^\aspil{P}_{\aspil{X},A,B}$,
means that any parameter of component \aspil{X} and label \aspil{P} must contain at least one level value (\aspil{I}) in the possible expression levels of \aspil{X}.
Indeed, the lower bound is 1, forcing at least one element in the parameter, but no upper bound is specified, allowing up to any number of answers.
The body (right-hand side) of the rule also checks for the existence of a parameter of \aspil{X} with label \aspil{P}, and constrains that the parametrization inference was not conclusive for the considered parameter (\aspil{not} stands for negation by failure: \aspil{not L} becomes true if \aspil{L} is not true).
Such a constraint gives multiple results as any set of \aspil{param} atoms satisfying the cardinality will lead to a new global set of answers.
In this way, we enumerate all possible parametrizations which respects the results of parameters
inference, but completely disregarding the notion of admissible parametrizations given in
\pref{ssec:admissible-K}.

We rely on integrity constraints to filter only admissible parametrizations.
An integrity constraint is a rule with no head, that makes an answer set unsatisfiable if its body turns out to be true.
Hence, if we suppose that:
\begin{itemize}
  \item the \aspil{less\_active(\ti{a}, \ti{p}, \ti{q})} atom means that $K^\ti{p}_{\ti{a},A,B}$ stands for a configuration with less activating regulators than $K^\ti{q}_{\ti{a},A',B'}$ (\ie $A \subset A'$),
  \item the \aspil{param\_inf(\ti{a}, \ti{p}, \ti{q})} atom means: $K^\ti{p}_{\ti{a},A,B} \leqsegm K^\ti{q}_{\ti{a},A',B'}$,
\end{itemize}
the monotonicity assumption is formulated as the following integrity constraint:
\asp{:- less\_active(X, P, Q), not param\_inf(X, P, Q).}
which removes all parametrization results where parameters $K^\aspil{P}_{\aspil{X},A,B}$ and $K^\aspil{Q}_{\aspil{X},A',B'}$ exist such that $A \subset A'$ and $K^\aspil{Q}_{\aspil{X},A',B'} \ltsegm K^\aspil{P}_{\aspil{X},A,B}$, thus violating the monotonicity assumption.
Of course, other assumptions can be formulated in the same way.

>>>>

\begin{property}[Extreme values assumption]
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{extreme values assumption} if and only if:
\label{pro:param_enum_extreme}
\begin{align*}
  \forall b \in \Gamma, \GRNreg{b} \neq \emptyset \Longrightarrow \exists \omega \subset \GRNreg{b}, 0 \in K_{b,\omega} \wedge \exists \omega' \subset \GRNreg{b}, l_b \in K_{b,\omega'}
\end{align*}
\end{property}

\begin{property}[Activity assumption]
\label{pro:param_enum_activity}
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{activity assumption} if and only if:
\begin{align*}
  \forall b \in \Gamma, \forall a \in \GRNreg{b}, \exists \omega \subset \GRNreg{b}, K_{b,\omega} \neq K_{b,\omega \cup \{ a \}}
\end{align*}
\end{property}

\begin{property}[Monotonicity assumption]
\label{pro:param_enum_monotonicity}
Let $\IG = (\Gamma, E)$ be an IG. A parametrization $K$ on $\IG$ satisfies the \emph{monotonicity assumption} if and only if:
\begin{align*}
  \forall b \in \Gamma,
  \forall A^+ \subset \{ a \in \Gamma \mid \GRNedge{a}{+}{t}{b} \in E_+ \}&,
  \forall A^- \subset \{ a \in \Gamma \mid \GRNedge{a}{-}{t}{b} \in E_- \},\\
  K_{b,\omega \cup A^-} & \leqsegm K_{b,\omega \cup A^+}
\end{align*}
\end{property}

@@

\newcommand{\atom}{\mathbf}
\newcommand{\la}{\leftarrow}
\newcommand{\var}[1]{#1}
\newcommand{\nota}{\neg}

Answer Set Programming (ASP) \cite{Baral03} has been chosen to address the enumeration of all admissible parametrizations.
The motivations are following:
\begin{itemize}
  \item ASP efficiently tackles the inherent complexity of the models we use, thus allowing a fast execution of the formal tools defined in this paper,
  \item it is convenient to enumerate a large set of possible answers,
  \item it allows to easily constrain the answers according to some properties.
\end{itemize}
We now synthesize some key points to better make the reader understand our ASP implementation with the enumeration example.

ASP is a logic programming paradigm based on a set of rules.
A rule takes the form of:
$$head \la body.$$
where $head$ is an atom and $body$ is a series of atoms or negations of atoms.
Solving an ASP program means finding an answer set, \ie a minimal set of atoms that fulfills all the rules.
Several answer sets can be a solution to the same ASP program and a solver can be directed to enumerate them all.

A rule with no $body$ part is called a fact, and its $head$ atom has to be present in all answer sets.
For instance, all information describing the studied model (PH and inferred IG \& parameters) are expressed in ASP using facts.
For functional purposes, we assign a unique label to every possible set of resources $\omega$ of a given component,
and in the following we note $K^p_{a,\omega}$ the parameter of component $a$ whose set of resources $\omega$ is assigned to the label $p$.
Then, to state the existence of a parameter $K^p_{a,\omega}$, an atom named $\atom{param\_label}$ is used in the following fact:
$$\atom{param\_label}(a, p).$$

Defining a set in ASP is equivalent to defining the rule for belonging to this set.
For example, we define an atom $param\_act$ that describes the set of active regulators of a given a parameter.
%of component \ti{a} and label \ti{p} (\ie the set $A$ of a parameter $K^\ti{p}_{\ti{a},A,B}$).
Describing the activators of $K^p_{a,\{b,c\}}$ gives:
\begin{align*}
  \atom{param\_act}(a, p, b).&&
  \atom{param\_act}(a, p, c).
\end{align*}
The absence of such a fact involving another potential regulator $d$ of $a$ with label $p$ indicates that $d$ is not in the set of resources related to this parameter.
\rewrite{Find another example.}

Rules allow more detailed declarations than facts as they have a body (right-hand part below) containing constraints and allowing to use variables, while facts only have a head (left-hand part).
For instance, in order to define the set of expression levels of a component, we declare:
$$\atom{component\_levels}(\var{X}, 0..\var{M}) \la \atom{component}(\var{X}, \var{M}).$$
where the atom $\atom{component}(X, M)$ stands for the existence of a component $X$ with a maximum level $M$.
Considering this declaration, any possible answer for the atom $\atom{component\_levels}$ will be found by binding all possible values of its two terms with all existing $\atom{component}$ facts:
the existence of an answer $\atom{component\_levels}(a, k)$ will depend on the existence of a term $a$, which is bound with $\var{X}$, and an integer $k$, constrained by: $0 \leq k \leq \var{M}$.

Cardinalities are convenient to enumerate all possible parametrizations by creating multiple answer sets.
A cardinality (denoted hereafter with curly brackets) gives any number of possible answers for some atoms between a lower and upper bounds.
For example,
\begin{align*}
  & 1\ \{ \atom{param}(\var{X}, \var{P}, \var{I}) : \atom{component\_levels}(\var{X}, \var{I}) \} \la \\
  & \qquad\qquad \atom{param\_label}(\var{X}, \var{P}), \nota \atom{infered\_param}(\var{X}, \var{P}).
\end{align*}
where $\atom{param}(\var{X}, \var{P}, \var{I})$ stands for: $\var{I} \in K^\var{P}_{\var{X},\omega}$,
means that any parameter of component $\var{X}$ and label $\var{P}$ must contain at least one level value ($\var{I}$) in the possible expression levels of $\var{X}$.
Indeed, the lower bound is $1$, forcing at least one element in the parameter, but no upper bound is specified, allowing up to any number of answers.
The body (right-hand side) of the rule also checks for the existence of a parameter of $\var{X}$ with label $\var{P}$,
and constrains that the parametrization inference was not conclusive for the considered parameter (“$\nota$” stands for negation by failure: $\nota \atom{L}$ becomes true if $\atom{L}$ is not true).
Such a constraint gives multiple results as any set of $\atom{param}$ atoms satisfying the cardinality will lead to a new global set of answers.
In this way, we enumerate all possible parametrizations which respects the results of parameters inference,
but completely disregarding the notion of admissible parametrizations given in \pref{ssec:admissible-K}.

We rely on integrity constraints to filter only admissible parametrizations.
An integrity constraint is a rule with no head, that makes an answer set unsatisfiable if its body turns out to be true.
Hence, if we suppose that:
\begin{itemize}
  \item the $\atom{less\_active}(a, p, q)$ atom means that $K^p_{a,\omega}$ stands for a configuration with less activating regulators than $K^q_{a,\omega'}$, % (\ie $A \subset A'$),
  \item the $\atom{param\_inf}(a, p, q)$ atom means: $K^p_{a,\omega} \leqsegm K^q_{a,\omega'}$,
\end{itemize}
then the monotonicity assumption (\pref{pro:param_enum_monotonicity}) is formulated as the following integrity constraint:
$$\la \atom{less\_active}(\var{X}, \var{P}, \var{Q}), \nota \atom{param\_inf}(\var{X}, \var{P}, \var{Q}).$$
which removes all parametrization results where parameters $K^\var{P}_{\var{X},\omega}$ and $K^\var{Q}_{\var{X},\omega'}$ exist such that $\var{X}$ is less activated by $\omega$ than $\omega'$ %$A \subset A'$
but $K^\var{Q}_{\var{X},\omega'} \ltsegm K^\var{P}_{\var{X},\omega}$,
thus violating the monotonicity assumption.
Of course, other assumptions can be formulated in the same way.
